### 2008-12-19 talk/ecoop2009/rainbow.tex
add one more para about virtuals, and start the example about promotion

@@

@@

A very different point of view on promotion is as a generalization of techniques
that already exist in dynamic compilers as found in modern virtual machines for
object-oriented language.  In this context feedback
techniques are crucial for good results.  The main goal is to
optimize and reduce the overhead of dynamic dispatching and indirect
invocation.  This is achieved with variations on the technique of
polymorphic inline caches \cite{hoelzle_optimizing_1991}: the dynamic lookups are cached and
the corresponding generated machine code contains chains of
compare-and-jump instructions which are modified at run-time.  These
techniques also allow the gathering of information to direct inlining for even
better optimization results.

@@

The new generated code is potentially different for each run-time value
seen.  This implies that the generated code needs to contain some sort
of updatable switch, which can pick the right code path based on the
run-time value.

\cfbolz{I think this example is confusing, it is unclear in which order things
happen how. I will try to come up with something different}.

\begin{itemize}
  %XXX remove mention of rainbow interp. but this needs to be rewritten anyway
  \item (compile time): the rainbow interpreter produces machine code until it
    hits a promotion point; e.g.::

    \begin{lstlisting}[language=C]
        y = hint(x, promote=True)
        return y+10
    \end{lstlisting}

  \item (compile time): at this point, it generates special machine code that when
    reached calls the JIT compiler again; the JIT compilation stops::

    \begin{lstlisting}[language=C]
        switch(y) {
            default: compile_more(y);
        }
    \end{lstlisting}

  \item (runtime): the machine code is executed; when it reaches a promotion
    point, it executes the special machine code we described in the previous
    point; the JIT compiler is invoked again;

  \item (compile time): now we finally know the exact value of our red variable,
    and we can promote it to green; suppose that the value of 'y' is 32::

    \begin{lstlisting}[language=C]
        switch(y) {
            32: return 42;
            default: compile_more(y);
        }
    \end{lstlisting}

    Note that the operation "y+10" has been constant-folded into "42", as it
    was a green operation.

  \item (runtime) the execution restart from the point it stopped, until a new
    unhandled promotion point is reached.
\end{itemize}

\subsection{Promotion as Applied to the TLC}

XXX maybe a tlc example can be used for the example above?

>>>>

\anto{Maybe we should talk about ``virtual instances'' and not structures,
  considering the context}

@@

Even if not shown in the example, \lstinline{stack} is not the only
virtualized object.  In particular the two objects created by
\lstinline{IntObj(0)} are also virtualized, and their fields are stored as
local variables as well.  Virtualizion of instances is important not only
because it avoids the allocation of unneeded temporary objects, but also
because it makes possible to optimize method calls on them, as the JIT
compiler knows their exact type in advance.

@@

A very different point of view on promotion is as a generalization of techniques
that already exist in dynamic compilers as found in modern virtual machines for
object-oriented language.  In this context feedback
techniques are crucial for good results.  The main goal is to
optimize and reduce the overhead of dynamic dispatching and indirect
invocation.  This is achieved with variations on the technique of
polymorphic inline caches \cite{hoelzle_optimizing_1991}: the dynamic lookups are cached and
the corresponding generated machine code contains chains of
compare-and-jump instructions which are modified at run-time.  These
techniques also allow the gathering of information to direct inlining for even
better optimization results. 
\anto{What about this: While traditional PICs are only applied to indirect
  calls, promotion is a more general operation that can be applied to any kind
  of value, including instances of user-defined classes or integer numbers}

@@

The new generated code is potentially different for each run-time value
seen.  This implies that the generated code needs to contain some sort
of updatable switch, or \emph{flexswitch}, which can pick the right code path based on the
run-time value.

Let us look again at the TLC example.  To ease the reading, figure
\ref{fig:tlc-main} showed a simplified version of TLC's main loop, which did
not include the hints.  The real implementation of the \lstinline{LT} opcode
is shown in figure \ref{fig:tlc-main-hints}.

\begin{figure}[h]
\begin{center}
\begin{lstlisting}[language=Python]
        elif opcode == LT:
            a, b = stack.pop(), stack.pop()
            hint(a, promote_class=True)
            hint(b, promote_class=True)
            stack.append(IntObj(b.lt(a)))
\end{lstlisting}
\caption{Usage of hints in TLC's main loop}
\label{fig:tlc-main-hints}
\end{center}
\end{figure}

By promoting the class of \lstlisting{a} and \lstlisting{b}, we tell the JIT
compiler not to generate code until it knows the exact RPython class of both.
